### 2013-08-12 inst/doc/msmsTests-Vignette.Rnw
Adds msmsTests to the repos and updates manifest.
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/msmsTests@79384 bc3139a8-67e5-0310-9ffc-ced21a209358

%\VignetteIndexEntry{msmsTests: post test filters to improve reproducibility}
%\VignetteDepends{msmsTests}
%\VignetteKeywords{multivariate, hplot}
%\VignettePackage{msmsTests}

\documentclass[12pt,a4paper,oneside]{article}
\usepackage{fullpage}
\usepackage{float}
\usepackage[section]{placeins}
\usepackage{enumerate}

\begin{document}
\SweaveOpts{keep.source=TRUE,concordance=TRUE}

\title{msmsTests package\\
       LC-MS/MS post test filters to improve reproducibility}
\author{Josep Gregori, Alex Sanchez, and Josep Villanueva\\
   Vall Hebron Institute of Oncology \&\\
   Statistics Dept. Barcelona University\\
  \texttt{jgregori@vhio.net}}
\date{July 2013}

\maketitle

\tableofcontents

\section{Introduction}

The \emph{omics} technologies are in general challenged by a high
dimensionality problem, whereby a high number of statistical tests are carried
out at the same time on very few samples. The high number of tests that are
carried out simultaneously requires a p-value adjustment to control the false
discovery rate (FDR). The low number of replicates, combined with a high
number of statistical tests easily leads to low reproducibility in the results. Shi et al. addressed this issue in a systematic manner using a large dataset
elaborated within the studies of the MicroArray Quality Control Consortium
(MAQC) \cite{shi2010}. The MAQC studied the possible sources of the limitations 
of microarray (MA) studies that precluded obtaining reproducible results across
laboratories and platforms. Their conclusions underscored the influence of
batch effects and the limitation introduced by using only the p-value criteria 
to get ordered lists of differentially expressed features as the main
weaknesses in MA studies \cite{shi2008} \cite{luo2010}.
This recommendation makes an intuitive appeal to the use of a post-test filter, excluding features that show low p-values but poor effect size or low signal strength, as a mean to guaranty a minimal level of reproducibility.

Label-free differential proteomics is based on comparing the expression of 
proteins between different biological conditions
\cite{mallik2010} \cite{neilson2011}. The expression of a protein by LC-MS/MS
may be obtained by MS peak intensity measures or by the number of spectral 
counts assigned to that protein in a LC-MS/MS run. Here we concentrate on 
spectral counts.

In proteomics reproducibility is also of great concern, and the use of such 
post-test filters may help notably \cite{gregori2013}.

\section{An example LC-MS/MS dataset}

The example dataset \cite{gregori2013} is the result of multiple spiking 
experiments, showing real LC-MS/MS data . Samples of 500 micrograms of a 
standard yeast lysate are spiked with 100, 200, 400 and 600fm of a complex 
mix of 48 equimolar human proteins (UPS1, Sigma-Aldrich). The number of 
technical replicates vary between 3 to 6.
\newline
The dataset consists in an instance of the \emph{MSnSet} class, defined in the 
MSnbase package \cite{Gatto2012}, a S4 class \cite{chambers} \cite{genolini}. 
This \emph{MSnSet} object contains
a spectral counts (SpC) matrix in the \emph{assayData}
slot, and a factor treatment in the \emph{phenoData} slot.
(See also the expressionSet vignette by vignette("ExpressionSetIntroduction",package="Biobase") \cite{gentleman})

<<echo=FALSE>>=
options(continue=" ")
@

<<setup.chunk, echo = TRUE>>=
library(msmsTests)
data(msms.spk)
msms.spk
dim(msms.spk)
head(pData(msms.spk))
table(pData(msms.spk)$treat)
@

Although the mix is equimolar the signal strength of each protein is markedly
different, allowing to cover the full range of SpC values, what makes it 
specially worth in this sort of experiments:

<<SpC.dist.chunk, echo = TRUE>>=
msms.spc <- exprs(msms.spk)
treat <- pData(msms.spk)
idx <- grep("HUMAN",rownames(msms.spc)) 
mSpC <- t( apply(msms.spc[idx,],1,function(x) tapply(x,treat,mean)) )
apply(mSpC,2,summary)
@

\section{Differential expression tests on spectral counts}

Spectral counts (SpC) is an integer measure which requires of tests suited to compare counts, as the GML methods based in the Poisson distribution, the 
negative-binomial, or the quasilikelihood \cite{agresti2002}

Generally speaking no test is superior to the other. They are just more or less
indicated in some cases. The Poisson regression requires the estimation of just one parameter, and is indicated when the number of replicates is low, two or
three. A drawback of the Poisson distribution is that its variance equals its 
mean, and it is not able to explain extra sources of variability a part of the
sampling. Quasi-likelihood is a distribution independent GLM, but requires of 
the estimation of two parameters and hence needs a higher number of replicates,
i.e. not less than four. The negative-binomial requires two parameters too, but 
we may use the implementation of this GLM in the edgeR package 
\cite{robinson2010} which uses an empirical Bayes method to share information
across features and may be employed with a restricted number of replicates; in
the worst case it limits with the Poisson solution.

\section{Poisson GLM regression example}

When using the Poisson distribution we implicitly accept a model not sensitive 
to biological variability \cite{agresti2002}. So it is just recommended in cases
where we have very few replicates, if any, and we do not expect a significant
biological variability between samples.

<<Chunk1.Poiss, echo=TRUE>>=
### Subset to the 200 and 600fm spikings
fl <- treat=="U200" | treat=="U600"
e <- msms.spk[,fl]
table(pData(e))
### Remove all zero rows
e <- pp.msms.data(e)
dim(e)

### Null and alternative model
null.f <- "y~1"
alt.f <- "y~treat"
### Normalizing condition
div <- apply(exprs(e),2,sum)
### Poisson GLM
pois.res <- msms.glm.pois(e,alt.f,null.f,div=div)
str(pois.res)

### DEPs on unadjusted p-values
sum(pois.res$p.value<=0.01)
### DEPs on multitest adjusted p-values
adjp <- p.adjust(pois.res$p.value,method="BH")
sum(adjp<=0.01)

### The top features
o <- order(pois.res$p.value)
head(pois.res[o,],20)

### How the UPS1 proteins get ordered in the list
grep("HUMAN",rownames(pois.res[o,]))

###  Truth table
nh <- length(grep("HUMAN",featureNames(e)))
ny <- length(grep("HUMAN",featureNames(e),invert=TRUE))
tp <- length(grep("HUMAN",rownames(pois.res)[adjp<=0.01]))
fp <- sum(adjp<=0.01)-tp
(tt.pois1 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))
@

\section{Quasi-likelihood GLM regression example}

The quasi-likelihood is a distribution free model that allows for 
overdispersion, and could be indicated where an appreciable source of
biological variability is expected. In this model, instead of 
specifying a probability distribution for the data, we just provide a 
relationship between mean and variance. This relationship takes the form of
a function, with a multiplicative factor known as the overdispersion, which
has to be estimated from the data \cite{agresti2002}. Its use in proteomics
has been documented by Li et al. (2010)\cite{li2010}.

<<Chunk2.QL, echo=TRUE>>=
### Quasi-likelihood GLM
ql.res <- msms.glm.qlll(e,alt.f,null.f,div=div)
str(ql.res)

### DEPs on unadjusted p-values
sum(ql.res$p.value<=0.01)
### DEPs on multitest adjusted p-values
adjp <- p.adjust(ql.res$p.value,method="BH")
sum(adjp<=0.01)

### The top features
o <- order(ql.res$p.value)
head(ql.res[o,],20)

### How the UPS1 proteins get ordered in the list
grep("HUMAN",rownames(ql.res[o,]))

###  Truth table
tp <- length(grep("HUMAN",rownames(ql.res)[adjp<=0.01]))
fp <- sum(adjp<=0.01)-tp
(tt.ql1 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))
@

\section{edgeR: negative binomial GLM regression example}

The negative-binomial provides another model that allows for overdispersion. The
implementation adopted in this package is entirely based in the solution
provided by the package edgeR \cite{robinson2010} which includes empirical
Bayes methods to share information among features, and thus may be employed
even when the number of replicates is as low as two. The negative-binomial is downward limited, when no overdispersion is observed, by the Poisson 
distribution.

<<Chunk3.edgeR, echo=TRUE>>=
### Negative-binomial
nb.res <- msms.edgeR(e,alt.f,null.f,div=div,fnm="treat")
str(nb.res)

### DEPs on unadjusted p-values
sum(nb.res$p.value<=0.01)
### DEPs on multitest adjusted p-values
adjp <- p.adjust(nb.res$p.value,method="BH")
sum(adjp<=0.01)

### The top features
o <- order(nb.res$p.value)
head(nb.res[o,],20)

### How the UPS1 proteins get ordered in the list
grep("HUMAN",rownames(nb.res[o,]))

###  Truth table
tp <- length(grep("HUMAN",rownames(nb.res)[adjp<=0.01]))
fp <- sum(adjp<=0.01)-tp
(tt.nb1 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))
@

\section{Reproducibility}

In the \emph{omics} field, reproducibility is of biggest concern. Very low 
p-values for a protein in an experiment are not enough to declare that protein 
as of interest as a biomarker. A good biomarker should give as well a 
reproducible signal, and posses a biologically significant effect size. 
According to our experience, a protein giving less than three counts in the 
most abundant condition results of poor reproducibility, to be declared as
statistically significant, experiment after experiment. On the other hand most
of the false positives in an spiking experiment show log fold changes below
1. These two observations \cite{gregori2013} allow to improve the results 
obtained in the previous
sections. The trick is to flag as relevant those proteins 
which have low p-values, high enough signal, and good effect size. In performing
this relevance filter we may even accept higher adjusted p-values than usual. 
This flagging is provided by the function \texttt{test.results}.

<<Chunk4.repro, echo=TRUE, tidy=TRUE>>=
###  Cut-off values for a relevant protein as biomarker
alpha.cut <- 0.05
SpC.cut <- 2
lFC.cut <- 1

###  Relevant proteins according to the Poisson GLM
pois.tbl <- test.results(pois.res,e,pData(e)$treat,"U600","U200",div,
                         alpha=alpha.cut,minSpC=SpC.cut,minLFC=lFC.cut,
                         method="BH")$tres
(pois.nms <- rownames(pois.tbl)[pois.tbl$DEP=="T"])

###  Truth table
ridx <- grep("HUMAN",pois.nms)
tp <- length(ridx)
fp <- length(pois.nms)-length(ridx)
(tt.pois2 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))

###  Relevant proteins according to the quasi-likelihood GLM
ql.tbl <- test.results(ql.res,e,pData(e)$treat,"U600","U200",div,
                    alpha=0.05,minSpC=2,minLFC=1,method="BH")$tres
(ql.nms <- rownames(ql.tbl)[ql.tbl$DEP=="T"])

###  Truth table
ridx <- grep("HUMAN",ql.nms)
tp <- length(ridx)
fp <- length(ql.nms)-length(ridx)
(tt.ql2 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))

###  Relevant proteins according to the negative-binomial GLM
nb.tbl <- test.results(nb.res,e,pData(e)$treat,"U600","U200",div,
                    alpha=0.05,minSpC=2,minLFC=1,method="BH")$tres
(nb.nms <- rownames(nb.tbl)[nb.tbl$DEP=="T"])

###  Truth table
ridx <- grep("HUMAN",nb.nms)
tp <- length(ridx)
fp <- length(nb.nms)-length(ridx)
(tt.nb2 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))
@

As you may see, without the post-test filter a relatively low
adjusted p-value cut-off is required to keep an acceptable number of false positives. The post-test filter allows to relax the p-value cut-off improving 
at the same time both the number of true positives and false positives.

<<Chunk5.summary, echo=FALSE, results=tex>>=
fl <- pois.tbl$adjp <= 0.05
sig <- sum(fl)
tp <- length(grep("HUMAN",rownames(pois.tbl)[fl]))
fp <- sig-tp
tt.pois0 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp)

fl <- ql.tbl$adjp <= 0.05
sig <- sum(fl)
tp <- length(grep("HUMAN",rownames(ql.tbl)[fl]))
fp <- sig-tp
tt.ql0 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp)

fl <- nb.tbl$adjp <= 0.05
sig <- sum(fl)
tp <- length(grep("HUMAN",rownames(nb.tbl)[fl]))
fp <- sig-tp
tt.nb0 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp)

library(xtable)
df <- data.frame(Test=character(9),Significance=numeric(9),
                 Filtered=character(9),
                 TP=integer(9),FP=integer(9),TN=integer(9),FN=integer(9),
                 stringsAsFactors=FALSE)
df$Test[1] <- df$Test[4] <- df$Test[7] <- "Poisson"
df$Test[2] <- df$Test[5] <- df$Test[8] <- "Quasi-likelihood"
df$Test[3] <- df$Test[6] <- df$Test[9] <- "Negative-binomial"
df$Significance[1:3] <- 0.05
df$Significance[4:6] <- 0.01
df$Significance[7:9] <- 0.05
df$Filtered[1:6] <- "No"
df$Filtered[7:9] <- "Yes"
df[1,4:7] <- tt.pois0
df[2,4:7] <- tt.ql0
df[3,4:7] <- tt.nb0
df[4,4:7] <- tt.pois1
df[5,4:7] <- tt.ql1
df[6,4:7] <- tt.nb1
df[7,4:7] <- tt.pois2
df[8,4:7] <- tt.ql2
df[9,4:7] <- tt.nb2
print(xtable(df,align=c("r","r","c","c","c","c","c","c"),
             caption=c("Truth tables"),
             display=c("d","s","f","s","d","d","d","d")),
      type="latex",hline.after=c(-1,0,3,6,9),include.rownames=FALSE)
@

\section{Other functions in the package}

A useful tool to visualize the global results of differential expression tests
is a table of accumulated frequencies of features by p-values in bins of
log fold changes. It may help in finding the most appropriate post-test filter
cut-off values in a given experiment.

<<Chunk6.other, echo=TRUE, tidy=TRUE>>=
### All features
pval.by.fc(ql.tbl$adjp,ql.tbl$LogFC)

### Filtering by minimal signal
fl <- ql.tbl$U600 > 2
pval.by.fc(ql.tbl$adjp[fl],ql.tbl$LogFC[fl])
@

Another usual tool is a volcanoplot with the ability to visualize the effect
of different post-test filter cut-off values.

\begin{figure}[H]
\centering
<<fig=TRUE, echo=TRUE, width=5, height=5>>=
par(mar=c(5,4,0.5,2)+0.1)
res.volcanoplot(ql.tbl,max.pval=0.05,min.LFC=1,maxx=3,maxy=NULL,
                            ylbls=3)
@
\caption{Volcanoplot}\label{volcano}
\end{figure}

\newpage

\begin{thebibliography}{11}
\bibitem{shi2010}
Shi, L. et al. \emph{MAQC Consortium: The MicroArray Quality Control (MAQC)-II
study of common practices for the development and validation of microarray-based
predictive models}. Nat Biotech 2010, 28, 827-838.
\bibitem{shi2008}
Shi, L. et al. \emph{The balance of reproducibility, sensitivity, and 
specificity of lists of differentially expressed genes in microarray studies}.
BMC Bioinformatics 2008, 9 Suppl 9, S10.
\bibitem{luo2010}
Luo, J. et al. \emph{A comparison of batch effect removal methods for 
enhancement of prediction performance using MAQC-II microarray gene expression
data}. The Pharmacogenomics Journal 2010, 10, 278-291.
\bibitem{mallik2010} Mallick P., Kuster B. \emph{Proteomics: a pragmatic 
perspective.} Nat Biotechnol 2010;28:695-709.
\bibitem{neilson2011}
Neilson K.A., Ali N.A., Muralidharan S., Mirzaei M., Mariani M.,
Assadourian G., et al. \emph{Less label, more free: approaches in
label-free quantitative mass spectrometry.} Proteomics
2011;11:535-53.
\bibitem{gregori2013}  
Gregori J., Villareal L., Sanchez A., Baselga J., 
Villanueva J.,  \emph{An Effect Size Filter Improves the Reproducibility in Spectral Counting-based Comparative Proteomics}. Journal of Proteomics 2013,  
\url{http://dx.doi.org/10.1016/j.jprot.2013.05.030}
\bibitem{Gatto2012} Laurent Gatto and Kathryn S. Lilley, MSnbase - an R/Bioconductor package for isobaric tagged mass spectrometry data visualization, processing and quantitation, Bioinformatics 28(2), 288-289 (2012).
\bibitem{chambers} Chambers J.M. \emph{Software for data analysis: programming
with R}, 2008 Springer
\bibitem{genolini} Genolini C. \emph{A (Not So) Short Introduction to S4} (2008)
\bibitem{gentleman} Falcon S., Morgan M., Gentleman R. \emph{An Introduction to
Bioconductor's ExpressionSet Class} (2007)
\bibitem{agresti2002}
Agresti A., \emph{Categorical Data Analysis},
Wiley-Interscience, Hoboken NJ, 2002
\bibitem{robinson2010}
Robinson MD, McCarthy DJ and Smyth GK (2010). \emph{edgeR: a Bioconductor 
package for differential expression analysis of digital gene expression data}. Bioinformatics 26, 139-140
\bibitem{li2010}
Li, M.; Gray, W.; Zhang, H.; Chung, C. H.; Billheimer, D.; Yarbrough, W.
G.; Liebler, D. C.; Shyr, Y.; Slebos, R. J. C. \emph{Comparative shotgun
proteomics using spectral count data and quasi-likelihood modeling}. 
J Proteome Res 2010, 9, 4295-4305
\bibitem{josep2012} Gregori J., Villareal L., Mendez O., Sanchez A.,
Baselga J., Villanueva J., \emph{Batch effects correction improves the 
sensitivity of significance tests in spectral counting-based comparative 
discovery proteomics}, Journal of Proteomics, 2012, 75, 3938-3951
\bibitem{BH} Benjamini, Y., and Hochberg, Y. (1995). \emph{Controlling the 
false discovery rate: a practical and powerful approach to multiple testing}. Journal of the Royal Statistical Society Series B, 57, 289-300.

\end{thebibliography}

\end{document}

### 2013-08-12 inst/doc/msmsTests-Vignette2.Rnw
Adds msmsTests to the repos and updates manifest.
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/msmsTests@79384 bc3139a8-67e5-0310-9ffc-ced21a209358

%\VignetteIndexEntry{msmsTests: controlling batch effects by blocking}
%\VignetteDepends{msmsTests}
%\VignetteKeywords{multivariate, hplot}
%\VignettePackage{msmsTests}

\documentclass[12pt,a4paper,oneside]{article}
\usepackage{fullpage}
\usepackage{float}
\usepackage[section]{placeins}
\usepackage{enumerate}

\begin{document}
\SweaveOpts{keep.source=TRUE,concordance=TRUE}

\title{msmsTests package\\
       Blocks design to compensate batch effects}
\author{Josep Gregori, Alex Sanchez, and Josep Villanueva\\
   Vall Hebron Institute of Oncology \&\\
   Statistics Dept. Barcelona University\\
  \texttt{jgregori@vhio.net}}
\date{July 2013}

\maketitle

\tableofcontents

\section{Introduction}

This vignette exemplifies the use of the packages msmsEDA and msmsTests
in discovering and correcting batch effects in label-free LC-MS/MS data
based on spectral counts. Label-free experiments are specially sensitive
to these effects as each condition has to be measured separately and may be
influenced by uncontrolled factors in a different extend.

\section{Dataset}

This dataset \cite{gregori2013} is the result of spiking experiments, 
showing real LC-MS/MS data. Samples of 500 micrograms of a standard 
yeast lysate are spiked with 200 and 600fm of a complex mix of 
48 equimolar human proteins (UPS1, Sigma-Aldrich).
The dataset comes with the package msmsEDA \cite{msmsEDA},
and was used to evidence batch effects by exploratory data analysis 
tools \cite{josep2012}.
The dataset consists in an instance of the \emph{MSnSet} class, defined in the 
MSnbase package \cite{Gatto2012}, a S4 class \cite{chambers} \cite{genolini}. 
This \emph{MSnSet} object contains
a spectral counts (SpC) matrix in the \emph{assayData}
slot, and factors treatment and batch in the \emph{phenoData} slot.
(See also the expressionSet vignette by vignette("ExpressionSetIntroduction",package="Biobase") \cite{gentleman})
<<echo=FALSE>>=
options(continue=" ")
@

<<SetupChunk, echo = TRUE>>=
library(msmsTests)
data(msms.dataset)
msms.dataset
msms.counts <- exprs(msms.dataset)
dim(msms.counts)
table(pData(msms.dataset)$treat,pData(msms.dataset)$batch)
@

Although the mix is equimolar the signal strength of each protein is markedly
different, allowing to cover a wide range of SpC values, what makes it 
specially worth in this sort of experiments:

<<DistChunk, echo = TRUE>>=
idx <- grep("HUMAN",featureNames(msms.dataset)) 
mSpC <- t( apply(msms.counts[idx,],1,function(x) 
                   tapply(x,pData(msms.dataset)$treat,mean)) )
apply(mSpC,2,summary)
@

\section{Batch effects}

Real life LC-MS/MS experiments use to be complicated enough 
to be able to get all required technical or biological replicates 
in a single batch run. Commonly a dataset collects results from
multiple batches. The batches may be influenced by factors 
which escape our control capacity, and typically these datasets
show the so known '\emph{batch effects}' when the runs where
obtained in different dates. The confounding
caused by these effects is easily evidenced by multidimensional
tools like Principal Components Analysis (PCA) or Hierarchical
Clustering (HC), when the samples cluster by batches instead
of by treatment \cite{josep2012} \cite{luo2010} .

<<PCAChunk, echo=TRUE, tidy=FALSE>>=
snms <- substr(as.character(pData(msms.dataset)$treat),1,2)
snms <- paste(snms,as.integer(pData(msms.dataset)$batch),sep=".")
smpl.pca <- counts.pca(msms.dataset,snms=snms)$pca
@

\begin{figure}[H]
\centering
<<fig=TRUE, echo=FALSE, width=6.5, height=4>>=
par(mar=c(4,4,0.5,2)+0.1)
facs <- data.frame(batch=pData(msms.dataset)$batch)
counts.pca(msms.dataset,facs=facs,snms=snms)
@
\caption{Principal Components Analysis}\label{pca}
\end{figure}

\section{Results on a single batch}

The next code shows the results obtained from the data of a single
batch with four replicates in each condition. The statistic test
used for differential expression is the quasi-likelihood GLM
\cite{agresti2002} \cite{li2010}, and
the p-values are adjusted with FDR control by the Benjamini-Hochberg
\cite{BH} method. The quality of the results is given by a truth table.

<<SubsetChunk, echo = TRUE>>=
###  Subset and pre-process dataset
fl <- pData(msms.dataset)$batch=="2502"
e <- msms.dataset[,fl]
e <- pp.msms.data(e)
### Null and alternative model
null.f <- "y~1"
alt.f <- "y~treat"
### Normalizing condition
counts <- exprs(e)
div <- apply(counts,2,sum)
### Quasi-likelihood GLM
ql.res <- msms.glm.qlll(e,alt.f,null.f,div=div)
### Adjust p-values with FDR control.
adjp <- p.adjust(ql.res$p.value,method="BH")
###  Truth table
nh <- length(grep("HUMAN",featureNames(e)))
ny <- length(grep("HUMAN",featureNames(e),invert=TRUE))
tp <- length(grep("HUMAN",rownames(ql.res)[adjp<=0.05]))
fp <- sum(adjp<=0.05)-tp
(tt.ql1 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))
@

These results may be polished by a post-test filter, so
that only relevant features are accepted as differentially expressed,
and the false positives are further restricted \cite{gregori2013}.

<<SubsetFChunk, echo = TRUE>>=
###  Post-test filter
ql.tbl <- test.results(ql.res,e,pData(e)$treat,"U600","U200",div,
                    alpha=0.05,minSpC=2,minLFC=1,method="BH")$tres
ql.nms <- rownames(ql.tbl)[ql.tbl$DEP=="T"]
###  Truth table
ridx <- grep("HUMAN",ql.nms)
tp <- length(ridx)
fp <- length(ql.nms)-length(ridx)
(tt.ql11 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))
@

\section{Results on the global dataset}

With a higher number of replicates the tests become more sensitive, and
a higher number of differentially expressed features may be identified.
The next code explores the full dataset, composed of two batches and
seven replicates of each condition. Again the quality of the results 
is given by a truth table.

<<GlobalChunk, echo = TRUE>>=
###  Pre-process dataset
gble <- pp.msms.data(msms.dataset)
### Null and alternative model
null.f <- "y~1"
alt.f <- "y~treat"
### Normalizing condition
div <- apply(exprs(gble),2,sum)
### Quasi-likelihood GLM
ql.res <- msms.glm.qlll(gble,alt.f,null.f,div=div)
### Adjust p-values with FDR control.
adjp <- p.adjust(ql.res$p.value,method="BH")
###  Truth table
nh <- length(grep("HUMAN",featureNames(gble)))
ny <- length(grep("HUMAN",featureNames(gble),invert=TRUE))
tp <- length(grep("HUMAN",rownames(ql.res)[adjp<=0.05]))
fp <- sum(adjp<=0.05)-tp
(tt.ql2 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))
@

Applying a post-test filter, as before, the results become:

<<GlobalFChunk, echo = TRUE>>=
###  Post-test filter
ql.tbl <- test.results(ql.res,gble,pData(gble)$treat,"U600","U200",div,
                    alpha=0.05,minSpC=2,minLFC=1,method="BH")$tres
ql.nms <- rownames(ql.tbl)[ql.tbl$DEP=="T"]
###  Truth table
ridx <- grep("HUMAN",ql.nms)
tp <- length(ridx)
fp <- length(ql.nms)-length(ridx)
(tt.ql22 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))
@

\section{Results on the global dataset with a blocking factor}

When the batches are balanced in the treatment conditions the
presence of confounding factors translates into bigger variance
and lower sensitivity. We may account for this extra variability 
by introducing the batches into the model, as a blocking factor.
The next code explores the corresponding results.

<<BlockChunk, echo = TRUE>>=
### Null and alternative model
null.f <- "y~batch"
alt.f <- "y~treat+batch"
### Quasi-likelihood GLM
ql.res <- msms.glm.qlll(gble,alt.f,null.f,div=div)
### Adjust p-values with FDR control.
adjp <- p.adjust(ql.res$p.value,method="BH")
###  Truth table
nh <- length(grep("HUMAN",featureNames(gble)))
ny <- length(grep("HUMAN",featureNames(gble),invert=TRUE))
tp <- length(grep("HUMAN",rownames(ql.res)[adjp<=0.05]))
fp <- sum(adjp<=0.05)-tp
(tt.ql3 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))
@

The correction improved the number of true positives, but significantly
increased the number of false positives. This may be polished by the
post-test filter to remove the non relevant features:

<<BlockFChunk, echo = TRUE>>=
###  Post-test filter
ql.tbl <- test.results(ql.res,gble,pData(gble)$treat,"U600","U200",div,
                    alpha=0.05,minSpC=2,minLFC=1,method="BH")$tres
ql.nms <- rownames(ql.tbl)[ql.tbl$DEP=="T"]
###  Truth table
ridx <- grep("HUMAN",ql.nms)
tp <- length(ridx)
fp <- length(ql.nms)-length(ridx)
(tt.ql33 <- data.frame(TP=tp,FP=fp,TN=ny-fp,FN=nh-tp))
@

\section{Comparison of results}

The following table collects the results obtained so far, where we see
how increasing the number of replicates we improve the sensitivity,
how the use of a post-test filter helps in restricting the number of
false positives, and how blocking helps to remove the extra variability
introduced by batch effects.

<<CompChunk, echo = FALSE, results=tex>>=
gbl.tt <- rbind(tt.ql1,tt.ql11,tt.ql2,tt.ql22,tt.ql3,tt.ql33)
rownames(gbl.tt) <- c("subset","subset filtered",
                      "global","global filtered",
                      "blocking","blocking filtered")		  
library(xtable)
print(xtable(gbl.tt,align=c("r","r","r","r","r"),
             caption=c("Truth tables"),
             display=c("s","d","d","d","d")),
      type="latex",hline.after=c(-1,0,2,4,6),include.rownames=TRUE)
@

\begin{figure}[H]
\centering
<<fig=TRUE, echo=FALSE, width=3, height=3.5>>=
par(mar=c(5, 3, 2, 2) + 0.1)
par(cex.axis=0.8,cex.lab=0.8)
rownames(gbl.tt) <- c("subset","subset\nfiltered",
                      "global","global\nfiltered",
                      "blocking","blocking\nfiltered")		  
barplot(t(data.matrix(gbl.tt[,1:2])),beside=TRUE,las=2,space=c(0,0.25),
        legend.text=c("TP","FP"),args.legend=list(x="topleft",cex=0.8,bty="n"))
@
\caption{Comparison of results}\label{barplot}
\end{figure}

\newpage

\begin{thebibliography}{11}
\bibitem{gregori2013}  
Gregori J., Villareal L., Sanchez A., Baselga J., 
Villanueva J.,  \emph{An Effect Size Filter Improves the Reproducibility in Spectral Counting-based Comparative Proteomics}. Journal of Proteomics 2013, 
\url{http://dx.doi.org/10.1016/j.jprot.2013.05.030}  
\bibitem{msmsEDA}
Gregori J., Sanchez A. and Villanueva J. (2013). \emph{msmsEDA: 
Exploratory Data Analysis of LC-MS/MS data by spectral counts}.
R package version 1.0.
\bibitem{josep2012} Gregori J., Villareal L., Mendez O., Sanchez A.,
Baselga J., Villanueva J., \emph{Batch effects correction improves the 
sensitivity of significance tests in spectral counting-based comparative 
discovery proteomics}, Journal of Proteomics, 2012, 75, 3938-3951
\bibitem{Gatto2012} Laurent Gatto and Kathryn S. Lilley, MSnbase - an R/Bioconductor package for isobaric tagged mass spectrometry data visualization, processing and quantitation, Bioinformatics 28(2), 288-289 (2012).
\bibitem{chambers} Chambers J.M. \emph{Software for data analysis: programming
with R}, 2008 Springer
\bibitem{genolini} Genolini C. \emph{A (Not So) Short Introduction to S4} (2008)
\bibitem{gentleman} Falcon S., Morgan M., Gentleman R. \emph{An Introduction to
Bioconductor's ExpressionSet Class} (2007)
\bibitem{luo2010}
Luo, J. et al. \emph{A comparison of batch effect removal methods for 
enhancement of prediction performance using MAQC-II microarray gene expression
data}. The Pharmacogenomics Journal 2010, 10, 278-291.
\bibitem{agresti2002}
Agresti A., \emph{Categorical Data Analysis},
Wiley-Interscience, Hoboken NJ, 2002
\bibitem{li2010}
Li, M.; Gray, W.; Zhang, H.; Chung, C. H.; Billheimer, D.; Yarbrough, W.
G.; Liebler, D. C.; Shyr, Y.; Slebos, R. J. C. \emph{Comparative shotgun
proteomics using spectral count data and quasi-likelihood modeling}. 
J Proteome Res 2010, 9, 4295-4305
\bibitem{BH} Benjamini, Y., and Hochberg, Y. (1995). \emph{Controlling the 
false discovery rate: a practical and powerful approach to multiple testing}. Journal of the Royal Statistical Society Series B, 57, 289-300.

\end{thebibliography}

\end{document}

### 2013-08-12 R/msmsTest-functions.R

@@ -1,0 +1,214 @@
+##################################################
+###  PROTEOMICS DIFFERENTIAL EXPRESSION TESTS  ###
+###       ON AN SPECTRAL COUNTS MATRIX         ###
+##################################################
+
+###  GLM Poisson test, given formula of Ha model 
+###   as: y~Treat  or  y~Treat+Batch   etc.
+###  against Ho model: y~1   or   y~Batch   etc.
+###  Returns a vector with estimated log fold change, residual 
+###   deviance, and unadjusted p-value.
+################################################################
+glm.pois <- function(count,form1,form0,facs,div)
+{ off=log(div)
+  data <- data.frame(y=round(count),facs,off,stringsAsFactors=TRUE)
+  gpois1a <- glm(as.formula(form1),offset=off,data=data,
+                  family=poisson(link=log))
+  logFC <- coef(gpois1a)[2]*log2(exp(1))
+  names(logFC) <- NULL
+  gpois1 <- glm(as.formula(form0),offset=off,data=data,
+                  family=poisson(link=log))
+  anovaPq <- data.frame(anova(gpois1, gpois1a, test="Chisq"))
+  p.val <- ifelse(anovaPq[2,4] < 0.1e-15, 1, anovaPq[2,5])
+  D <- anovaPq$Deviance[2]
+  c(logFC=logFC,D=D,p.val=p.val)
+}
+
+###  Model multifactor, regressio de Poisson
+##############################################
+msms.glm.pois <- function(msnset,form1,form0,facs=NULL,div=NULL)
+{ data <- exprs(msnset)
+  if(is.null(facs)) facs <- as.matrix(pData(msnset))
+  if(is.null(div)) div <- rep(1,ncol(data))
+  Np <- nrow(data)
+  result <- data.frame(LogFC=numeric(Np),D=numeric(Np),
+                       p.value= numeric(Np))
+  for(i in 1:Np) 
+    result[i,] <- glm.pois(data[i,],form1,form0,facs,div)
+  rownames(result) <- rownames(data)
+  result
+}
+
+
+###  Quasi-likelihood test, given formula of Ha model 
+###   as: y~Treat  or  y~Treat+Batch   etc.
[…]

### 2013-09-27 inst/doc/msmsTests-Vignette.Rnw
use hyperref LaTeX package so vignette will build on Linux and Mac
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/msmsTests@79450 bc3139a8-67e5-0310-9ffc-ced21a209358
In function test.results(), DEP column changed to logical
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/msmsTests@80847 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
    Statistics Dept. Barcelona University\\
-  \texttt{jgregori@vhio.net}}
+  \texttt{josep.gregori@gmail.com}}
 \date{July 2013}
@@
                          method="BH")$tres
-(pois.nms <- rownames(pois.tbl)[pois.tbl$DEP=="T"])
+(pois.nms <- rownames(pois.tbl)[pois.tbl$DEP])
 
@@
                     alpha=0.05,minSpC=2,minLFC=1,method="BH")$tres
-(ql.nms <- rownames(ql.tbl)[ql.tbl$DEP=="T"])
+(ql.nms <- rownames(ql.tbl)[ql.tbl$DEP])
 
@@
                     alpha=0.05,minSpC=2,minLFC=1,method="BH")$tres
-(nb.nms <- rownames(nb.tbl)[nb.tbl$DEP=="T"])
+(nb.nms <- rownames(nb.tbl)[nb.tbl$DEP])
 

### 2013-09-27 inst/doc/msmsTests-Vignette2.Rnw
add hyperref package to 2nd vignette as well
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/msmsTests@79485 bc3139a8-67e5-0310-9ffc-ced21a209358
In function test.results(), DEP column changed to logical
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/msmsTests@80847 bc3139a8-67e5-0310-9ffc-ced21a209358

@@
                     alpha=0.05,minSpC=2,minLFC=1,method="BH")$tres
-ql.nms <- rownames(ql.tbl)[ql.tbl$DEP=="T"]
+ql.nms <- rownames(ql.tbl)[ql.tbl$DEP]
 ###  Truth table

### 2013-09-27 R/msmsTest-functions.R

@@ -1,6 +1,8 @@
 ##################################################
 ###  PROTEOMICS DIFFERENTIAL EXPRESSION TESTS  ###
 ###       ON AN SPECTRAL COUNTS MATRIX         ###
+###                by J. Gregori               ###
+###              3rd August, 2013              ###
 ##################################################
 
 ###  GLM Poisson test, given formula of Ha model 
@@ -147,7 +149,7 @@ test.results <- function(test,msnset,gpf,gp1,gp2,div=NULL,
   mx.mean <- apply(mcpg[,c(iA,iB)],1,max)
   flags <- (adjp < alpha) & (mx.mean >= minSpC) & 
              abs(test[nms,"LogFC"]) >= minLFC
-  rm$DEP <- ifelse(flags,"T","F")
+  rm$DEP <- flags
   
   n <- sum(flags)
   nms <- nms[flags]
@@ -203,7 +205,7 @@ res.volcanoplot <- function(tres,max.pval=0.05,min.LFC=1,maxx=3,maxy=10,
   pflags <- tres$adjp <= max.pval & abs(tres$LogFC) >= min.LFC
   points(tres$LogFC[pflags],-log10(tres$adjp[pflags]),pch=21,bg="blue",
          cex=0.7)
-  fflags <- pflags & tres$DEP=="F"
+  fflags <- pflags & !tres$DEP
   points(tres$LogFC[fflags],-log10(tres$adjp[fflags]),pch=21,bg="red",
          cex=0.7)
   fflags <- pflags & -log10(tres$adjp) > ylbls

### 2013-10-01 R/msmsTest-functions.R
Updating vignettes
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/msmsTests@81013 bc3139a8-67e5-0310-9ffc-ced21a209358

@@ -1,8 +1,8 @@
 ##################################################
 ###  PROTEOMICS DIFFERENTIAL EXPRESSION TESTS  ###
-###       ON AN SPECTRAL COUNTS MATRIX         ###
+###         ON SPECTRAL COUNTS MATRIX          ###
 ###                by J. Gregori               ###
-###              3rd August, 2013              ###
+###               September, 2013              ###
 ##################################################
 
 ###  GLM Poisson test, given formula of Ha model 
@@ -186,7 +186,6 @@ pval.by.fc <- function(pvals,lfc)
 }
 
 
-
 ####################
 ###  Volcano plot
 ####################
@@ -213,4 +212,3 @@ res.volcanoplot <- function(tres,max.pval=0.05,min.LFC=1,maxx=3,maxy=10,
     text(tres$LogFC[fflags],-log10(tres$adjp[fflags])+maxy/25,cex=0.7,
          font=2,which(fflags),col="blue")
 }
-
